---
title: "QC + DE comparison RSEM & conservative counts"
author: "Marta Perez Alcantara"
date: "5 de septiembre de 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Initial exploration and QC of count data

In this document I will explore the two versions of counts data I have: conservative counts and RSEM. I will then perform the filtering and QC of both, and the differential expression analysis.

```{r setup, echo=FALSE}

# Load necessary libraries

library(readr) # fast reading of large files
library(edgeR)
library(biomaRt)  # for gene information
library(pheatmap)
library(RColorBrewer)
library(org.Hs.eg.db)
library(limma)
library(ggbiplot)
library(ggrepel) # provides geoms for ggplot2 to repel overlapping text labels.
library(ggplot2)
library(ggfortify)


# Working directory:
setwd("/Users/Marta/Documents/WTCHG/DPhil/Data/Diff_v2")

my_files <- list.files("/Users/Marta/Documents/WTCHG/DPhil/Data/Diff_v2",pattern=".results")   # creates list of files to import, containing gene-level count data (RSEM)
# pattern chooses only files that include ".results". That way I exclude directories I have under my folder "Diff_v2".


donors= c("Ad2.1","Ad3.1","Neo1.1")  # data comes from three donors

stage= c("iPSC", "DE", "PGT", "PFG", "PE", "EP","ENstage6", "ENstage7") # 8 stages




```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
